Simulating an Isolated Dwarf Galaxy with Energetic Feedback and Chemical Yields from Individual Stars

### 2018-03-16 main.tex
most of 6 finished

{\bf New (response to KVJ comment on comparability):} As noted in Section~\ref{sec:IC}, our galaxy model is not intended to directly reproduce the observed properties of Local Group UFD's. Notably, our initial conditions neglect a pre-existing stellar population and are only followed for 500~Myr, a fraction of the age of z = 0 dwarf galaxies. However, we can still place our model in context with observations using simple comparisons to the star formation rate (Section~\ref{sec:gas_sf}), molecular gas (Section~\ref{sec:molecular gas content}), and metal retention fraction (Section~\ref{sec:obs_metals}) properties of observed dwarf galaxies. We show that these properties are broadly consistent with observations.

@@

The observational sample of isolated, gaseous, low mass dwarf galaxies is limited compared to more massive galaxies, but has improved substantially with recent blind and targeted \ion{H}{1} surveys \citep[e.g.][]{Giovanelli2005, Geha2006, Geha2012, Walter2008, Cannon2011, Haynes2011, Hunter2012, Bradford2015, James2015, Tollerud2015, Sand2015, Wang2017}. However, the sample of isolated, gaseous dwarf galaxies with $M_{*} < 10^{7}$~M$_{\odot}$ remains small. In Figure ~\ref{fig:KS} we show where our galaxy lies relative to the observed Kennicutt-Schmidt (KS) relation and extended Schmidt law (ESL)  for low mass galaxies. In both diagrams, our simulations are given by the colored points, sampled every 1 Myr throughout the entire simulation.

Although simple to measure in simulations, these quantities are challenging to directly compare to observations. sWe have attempted to make a reasonable analog to how $\Sigma_{\rm sfr}$ and $\Sigma_{\rm gas}$ are measured observationally for low mass dwarfs \citet[see ][]{Roychowdhury2014}. We define $\Sigma_{\rm sfr} = \dot{M}_{*,10} / A_{*,10}$, where $\dot{M}_{*,10}$ is the SFR measured over the preceding 10~Myr, and $A_{*,10}$ is the area of the disk within the radius of the outermost star formed within the previous 10 Myr. Likewise, $\Sigma_{\rm gas} = M_{\rm gas,10} / A_{*,10}$, where M$_{\rm gas,10}$ is the total gas mass within this defined disk. However, the total gas content cannot be determined observationally. To match this limitation, we follow \cite{Roychowdhury2014} and take $\Sigma_{\rm gas, obs} = 1.34 \times \Sigma_{\rm HI}$, where the factor 1.34 attempts to account for He. We note that there is generally no correction made for any possible H$_{\rm 2}$ or \ion{H}{2} content. As shown in Section~\ref{sec:molecular gas content}, these components may be significant.

\begin{figure*}
\centering
\includegraphics[width=0.475\linewidth]{all_gas_schmidt_law_evolution}
\includegraphics[width=0.475\linewidth]{all_gass_efficiency_evolution}
\caption{The KS law (left) and ESL (right) relationships for our galaxy as measured every megayear, plotted as points colored by time, with dark / purple early and light / green late. See text for details of the calculation.
Recent observations from the S\ion{H}{1}ELD sample \citep{Teich2016} are plotted as black points with error bars. On the left, we also give the best fit line to galaxies from the FIGGS sample from \cite{Roychowdhury2014}, and on the right we also show the best fit line and 1 $\sigma$ errors from \cite{Shi2011}.   There is no clear correlation with time in this diagram.}
\label{fig:KS}
\end{figure*}

@@

In constructing our galaxy model, we employed \textit{no} tuning of the underlying physics, adopting only canonical values for any available free parameters. It is thus non-trivial that our galaxy should oscillate about the median relationships in Fig.~\ref{fig:KS}, and signifies a proper accounting of the relevant physics governing gas and star forming properties in our galaxy. This result is consistent with galactic evolution simulations run at high resolution with a detailed accounting of stellar feedback physics \citep[see ][ and references therein]{NaabOstriker2017} and with the demonstration by \citet{Li2005} that the KS law can be reproduced by gravitation acting on an isothermal disk. %AE, which was argued to demonstrate that this represents the basic physics at work.

@@

The molecular gas content of low mass dwarf galaxies is generally assumed
to be small, but is not well constrained by either theory or observations. Assuming the relationship in \citet{Leroy2013} and \citet{Momose2013}, \citet{Roychowdhury2014} finds typical molecular gas mass fractions can be anywhere from $f_{\rm H_2} = 0.05$ to $f_{\rm H_2} = 0.5$; a significant range in values. As discussed in Section~\ref{sec:properties}, our galaxy has $f_{H_2}$ of 0.17--0.20, within this range, but significantly exceeding the few percent often found in models. Ignoring the H$_2$ component of our galaxies would lead to an average shift of 0.24 dex to the right in the KS relationship, and 0.24 dex right and down in the ESL. This does not result in a significant disagreement with the observed relationships, but would constitute a quantitative change to the observed fits if our $f_{\rm H_2}$ is robust for observed low mass, low metallicity dwarf galaxies.

Interestingly, in this low-dust regime, our H$_2$ formation is dominated by the gas-phase reaction H$^-$~+~H~$\rightarrow$~H$_2$~+~e$^{-}$. In our model, H$^{-}$ is produced solely through the reaction H~+~e$^{-} \rightarrow$ H$^{-}$~+~$\gamma$. Prior to the establishment of a local stellar radiation field, e$^{-}$ production is dominated by the ionization of H and He from the UV background. This has the seemingly counterintuitive effect that turning off the UV background actually decreases the formation rate of H$_2$, as confirmed in test simulations. In a test simulation ignoring all H$^{-}$ reactions, we find H$_{2}$ mass fractions on order of  $\sim 5\%$ by first star formation, a factor of 3 reduction from our fiducial model.

@@

It is interesting to consider whether ejected metals in our model reside in the CGM or have been ejected into the intergalactic medium. While this cannot be observationally determined for dwarf galaxies at this mass, cosmological simulations of a $M_{\rm vir} = 10^{10}$ M$_{\odot}$ galaxy show that, by redshift zero 40\% is ejected from the galaxy's halo \citep{Angles-Alcazar2017}. We find 13.7\% of all metals lie within 1 kpc of the center of our galaxy, 26.5\% within 0.25 $R_{\rm vir}$ (or 6.85 kpc), and 38.6\% outside $R_{\rm vir}$. This is consistent with previous works, but this is again an evolving quantity. In addition, the amount of gas that escapes the virial radius is certainly sensitive to the details of gas accretion from the IGM on this galaxy, which we cannot capture in this model.

@@

\subsubsection{Full Stellar Wind Model}
\label{sec:stellar winds discussion}
Our stellar wind model significantly reduces the injected wind velocity from $\sim$1000~km~s$^{-1}$ to 20~km~s$^{-1}$. Although our model is capable of generating realistic stellar winds with temperature profiles and radial evolution comparable to those expected from analytic calculations \citep{Weaver1977}, the hot, fast winds place a near constant and severe constraint on the Courant time step that renders $\gtrsim$100~Myr simulations impractical. When considered in isolation, stellar winds are an important source of pre-SN feedback and do dramatically influence the dynamical evolution on molecular cloud and galaxy scales \citep{Dale2008,Peters2017,Gatto2017}. However, when considered together with ionizing radiation, stellar winds contain less total energy \citep{Agertz2013} and do not seem to have a significant dynamical influence in either idealized simulations \citep{Geen2015} or  individual giant molecular clouds \citep{Dale2014}. They are even less relevant in the low-metallicity regime studied here, as stellar winds become weaker with decreasing metallicity \citep{Puls2000, Vink2005}. Although they likely have minimal dynamical importance at resolutions where peak densities are not anyway high enough to trap ionization fronts, a full model of stellar winds may affect detailed ISM properties and metal mixing, warranting a closer examination in a future work.

@@

We do note that an isotropic, two-fluid model for cosmic ray feedback exists in  \textsc{Enzo} \citep{SalemBryan2014,Salem2015} and has been well tested. Mechanically, including cosmic ray feedback in our model is trivial. However, the cosmic ray population, their diffusion coefficient, and the magnetic field structure of the lowest mass dwarf galaxies each have significant enough uncertainties to warrant reserving their full inclusion into our model to a later work.

@@

Roughly half of massive stars live in binary pairs \citep{Sana2013}. Their interactions, primarily through mass transfer, can significantly alter their radiation properties and lifetimes. This can change both how much and how long these stars emit ionizing radiation, an important source of stellar feedback, and where and when these stars explode as SNe. This effect is rarely accounted for in galaxy evolution models, which are commonly based on calculations of single star evolution (e.g. STARBURST99), but could be significant. For example, \citet{Zapartas2017} finds that binarity extends the timescales over which core collapse SNe occur from a given star formation event, from a maximum time of $\sim$ 50~Myr to $\sim$~200~Myr. Although they find only $\sim 15\%$ of core collapse SNe explode after 50~Myr, this could still be an important effect. Properly accounting for the delay times due to variations in individual star lifetimes has already been shown to change the significance of feedback and influence galaxy metallicity properties \citep{Kimm2015}. Extending the lifetimes of these stars could increase the role of radiation feedback, however this may be less important as these additional photons are more likely to escape the galaxy \citep[e.g.][]{Ma2016-binary}.

@@

\section{Conclusion}
\label{sec:conclusion}
We have developed a new method for simulating galaxy evolution with detailed feedback and chemical enrichment. For the first time on galaxy scales, we simultaneously model multi-channel stellar feedback in detail, accounting for stellar winds, core collapse and Type Ia SNe, ionizing radiation followed through radiative transfer, photoelectric heating, and Lyman-Werner radiation using individual star particles. This, coupled with the detailed chemistry and heating/cooling physics followed with \textsc{Grackle}, allows us to capture realistic galaxy evolution in detail. In this work, we apply these methods to simulate the evolution of an isolated, low-mass, dwarf galaxy modeled after the $z=0$ properties of the Leo P ultrafaint dwarf galaxy. We present an overview of the properties of this simulation 
in this work.
% AE: mention companion paper here if publish at same time

For our simulated dwarf galaxy, we find:
\begin{enumerate}
\item Multi-channel feedback is effective in this galaxy in regulating star formation to occur at a rate consistent with the KS relationship and the extended Schmidt law in observed galaxies. See Figures~\ref{fig:sfr_mass_evolution} and \ref{fig:KS}.

@@

\item A thin, yet resolved, gas scale height ($\sim 50$~pc) beyond our stellar radius, with larger scale heights ($\sim 100$--200~pc) driven by feedback interior to the stellar radius. This is comparable to the resolution limit of the diffuse \ion{H}{1} in observed, gaseous ultrafaint dwarfs. See Figure~\ref{fig:scale_height}.

\item The ISRF of our galaxy is highly varying in time, by orders of magnitude, though it is relatively uniform in space at any given time. It is unclear how important these fluctuations are as a source of feedback, or if the affect can be approximated with a time averaged radial profile. See Figures \ref{fig:ISRF} and \ref{fig:ISRF_2D}.

\item (\textbf{Possibly remove depending on follow up}) Formation of a significant amount of H$_2$ through gas phase reactions with H$^-$. This dense gas is more resistive to feedback driven outflows than the less dense gas in the galaxy. This leads to high molecular gas fractions, with an H$_2$ mass comparable to that of \ion{H}{1} by the end of the simulation. This result requires further investigation to determine if this behavior is universally true of ultrafaint dwarf galaxies. See Figure~\ref{fig:ISM_evolution}.

>>>>

%mm3 {\bf New (response to KVJ comment on comparability):} 
As noted in Section~\ref{sec:IC}, our galaxy model is not intended to directly reproduce the observed properties of Local Group ultra-faint dwarfs. Notably, our initial conditions neglect a pre-existing stellar population and are only followed for 500~Myr, a fraction of the age of $z = 0$ dwarf galaxies. However, we can still place our model in context with observations using simple comparisons to the star formation rate (Section~\ref{sec:gas_sf}), molecular gas (Section~\ref{sec:molecular gas content}), and metal retention fraction (Section~\ref{sec:obs_metals}) properties of observed dwarf galaxies. We show that these properties are broadly consistent with observations.

@@

The observational sample of isolated, gaseous, low mass dwarf galaxies is limited compared to more massive galaxies, but has improved substantially with recent blind and targeted \ion{H}{1} surveys \citep[e.g.][]{Giovanelli2005, Geha2006, Geha2012, Walter2008, Cannon2011, Haynes2011, Hunter2012, Bradford2015, James2015, Tollerud2015, Sand2015, Wang2017}. However, the sample of isolated, gaseous dwarf galaxies with $M_{*} < 10^{7}$~M$_{\odot}$ remains small. In Figure ~\ref{fig:KS} we show where our galaxy lies relative to the observed Kennicutt-Schmidt relation and extended Schmidt law for low mass galaxies. In both diagrams, our simulations are given by the colored points, sampled every megayear throughout the entire simulation.

Although simple to measure in simulations, these quantities are challenging to directly compare to observations. We have attempted to make a reasonable analog to how $\Sigma_{\rm sfr}$ and $\Sigma_{\rm gas}$ are measured observationally for low mass dwarfs \citet[see ][]{Roychowdhury2014}. We define $\Sigma_{\rm sfr} = \dot{M}_{*,10} / A_{*,10}$, where $\dot{M}_{*,10}$ is the SFR measured over the preceding 10~Myr, and $A_{*,10}$ is the area of the disk within the radius of the outermost star formed within the previous 10 Myr. Likewise, $\Sigma_{\rm gas} = M_{\rm gas,10} / A_{*,10}$, where M$_{\rm gas,10}$ is the total gas mass within this defined disk. However, the total gas content cannot be determined observationally. To match this limitation, we follow \cite{Roychowdhury2014} and take $\Sigma_{\rm gas, obs} = 1.34 \times \Sigma_{\rm HI}$, where the factor 1.34 attempts to account for He. We note that there is generally no correction made for any possible H$_{\rm 2}$ or \ion{H}{2} content. As shown in Section~\ref{sec:molecular gas content}, these components may be significant.

\begin{figure*}
\centering
\includegraphics[width=0.475\linewidth]{all_gas_schmidt_law_evolution}
\includegraphics[width=0.475\linewidth]{all_gass_efficiency_evolution}
\caption{The Kennicutt-Schmidt law (left) and extended Schmidt law (right) relationships for our galaxy as measured every megayear, plotted as points colored by time, with dark / purple early and light / green late. See text for details of the calculation.
Recent observations from the S\ion{H}{1}ELD sample \citep{Teich2016} are plotted as black points with error bars. On the left, we also give the best fit line to galaxies from the FIGGS sample from \cite{Roychowdhury2014}, and on the right we also show the best fit line and 1 $\sigma$ errors from \cite{Shi2011}.   There is no clear correlation with time in this diagram.}
\label{fig:KS}
\end{figure*}

@@

In constructing our galaxy model, we employed \textit{no} tuning of the underlying physics, adopting only canonical values for any available free parameters. It is thus non-trivial that our galaxy should oscillate about the median relationships in Fig.~\ref{fig:KS}, and signifies a proper accounting of the relevant physics governing gas and star forming properties in our galaxy. This result is consistent with galactic evolution simulations run at high resolution with a detailed accounting of stellar feedback physics \citep[see ][ and references therein]{NaabOstriker2017} and with the demonstration by \citet{Li2005} that the Kennicutt-Schmidt law can be reproduced by gravitation acting on an isothermal disk. %AE, which was argued to demonstrate that this represents the basic physics at work.

@@

The molecular gas content of low-mass dwarf galaxies is generally assumed
to be small, but is not well constrained by either theory or observations. Assuming the relationship in \citet{Leroy2013} and \citet{Momose2013}, \citet{Roychowdhury2014} finds typical molecular gas mass fractions can be anywhere from $f_{\rm H_2} = 0.05$ to $f_{\rm H_2} = 0.5$; a significant range in values. As discussed in Section~\ref{sec:properties}, our galaxy has $f_{H_2}$ of 0.17--0.20, within this range, but significantly exceeding the few percent often found in models. Ignoring the H$_2$ component of our galaxies would lead to an average shift of 0.24 dex to the right in the Kennicutt-Schmidt law, and 0.24 dex right and down in the extended Schmidt law. This does not result in a significant disagreement with the observed relationships, but would constitute a quantitative change to the observed fits if our $f_{\rm H_2}$ is robust for observed low-mass, low-metallicity dwarf galaxies.

Interestingly, in this low-dust regime, our H$_2$ formation is dominated by the gas-phase reaction H$^-$~+~H~$\rightarrow$~H$_2$~+~e$^{-}$. In our model, H$^{-}$ is produced solely through the reaction H~+~e$^{-} \rightarrow$ H$^{-}$~+~$\gamma$. Prior to the establishment of a local stellar radiation field, e$^{-}$ production is dominated by the ionization of H and He from the UV background. This has the seemingly counterintuitive effect that turning off the UV background actually decreases the formation rate of H$_2$, as confirmed in test simulations. In a test simulation ignoring all H$^{-}$ reactions, we find H$_{2}$ mass fractions on order of  $\sim 5\%$ by first star formation, a factor of 3 reduction from our fiducial model.
%mm3 
    {\bf [Is there any treatment in the model of the visible light radiation field, which is sufficient to destroy H$^-$?]}

@@

It is interesting to consider whether ejected metals in our model reside in the CGM or have been ejected into the intergalactic medium. While this cannot be observationally determined for dwarf galaxies at this mass, cosmological simulations of an $M_{\rm vir} = 10^{10}$ M$_{\odot}$ galaxy show that, by redshift zero 40\% is ejected from the galaxy's halo \citep{Angles-Alcazar2017}. We find 13.7\% of all metals lie within 1 kpc of the center of our galaxy, 26.5\% within 0.25 $R_{\rm vir}$ (or 6.85 kpc), and 38.6\% outside $R_{\rm vir}$. This is consistent with previous works, but this is again an evolving quantity. In addition, the amount of gas that escapes the virial radius is certainly sensitive to the details of gas accretion from the IGM on this galaxy, which we cannot capture in this model.

@@

%mm3 \subsubsection{Full Stellar Wind Model}
    \subsubsection{Massive Stellar Wind Energy}
\label{sec:stellar winds discussion}
Our 
%mm3
   massive
stellar wind model 
%mm3 significantly 
     drastically 
reduces the injected wind velocity from $\sim$1000~km~s$^{-1}$ to 20~km~s$^{-1}$. Although our 
%mm3 model is 
   algorithm is entirely 
capable of generating realistic stellar winds with
%mm3 temperature profiles and radial evolution comparable to those expected from analytic calculations \citep{Weaver1977}, the hot, [the winds are not hot, only the post-shock gas; Weaver et al. calculated the bubble structure resulting from winds moving at observed speeds] 
     velocities comparable to those observed, such
fast winds place a near constant and severe constraint on the Courant time step that renders $\gtrsim$100~Myr simulations impractical. When considered in isolation, stellar winds are an important source of pre-SN feedback and 
%mm3 do 
   can
dramatically influence dynamical evolution on molecular cloud and galaxy scales \citep{Dale2008,Peters2017,Gatto2017}. However, when considered together with ionizing radiation, stellar winds contain less total energy \citep{Agertz2013} and do not seem to have a significant dynamical influence in either idealized simulations \citep{Geen2015} or  individual giant molecular clouds \citep{Dale2014},
%mm3 [insert text]
    unless densities near the ionizing source are high enough to trap the  \ion{H}{2} region in the source cell. In that case, they can clear out a cavity to allow initial establishment of the \ion{H}{2} region.

% [end of insertion]
They are even less relevant in the low-metallicity regime studied here, as stellar winds become weaker with decreasing metallicity \citep{Puls2000, Vink2005}. Although they likely have minimal dynamical importance at resolutions where peak densities are not anyway high enough to trap ionization fronts, a full model of stellar winds may affect detailed ISM properties and metal mixing, warranting closer examination in future work.

@@

We do note that an isotropic, two-fluid model for cosmic ray feedback exists in  \textsc{Enzo} \citep{SalemBryan2014,Salem2015} and has been well tested. Mechanically, including
%mm3 
   this relatively simple treatment of 
cosmic ray feedback in our model is trivial. However, the cosmic ray population, their diffusion coefficient, and the magnetic field structure of the lowest mass dwarf galaxies each have significant enough uncertainties to warrant reserving their full inclusion into our model to later work.

@@

Roughly half of massive stars live in binary pairs \citep{Sana2013}. Their interactions, primarily through mass transfer, can significantly alter their radiation properties and lifetimes. This can change both how much and how long these stars emit ionizing radiation, an important source of stellar feedback, and where and when these stars explode as SNe. This effect could be significant, but is rarely accounted for in galaxy evolution models, which are commonly based on calculations of single star evolution (e.g. STARBURST99). For example, \citet{Zapartas2017} finds that binarity extends the timescales over which core collapse SNe occur from a given star formation event, from a maximum time of $\sim$ 50~Myr to $\sim$~200~Myr. Although they find only $\sim 15\%$ of core collapse SNe explode after 50~Myr, this could still be an important effect. Properly accounting for the delay times due to variations in individual star lifetimes has already been shown to change the significance of feedback and influence galaxy metallicity properties \citep{Kimm2015}. Extending the lifetimes of these stars could increase the role of radiation feedback, however this may be less important as these additional photons are more likely to escape the galaxy \citep[e.g.][]{Ma2016-binary}.

@@

\section{Conclusion}
\label{sec:conclusion}
We have developed a new method for simulating galaxy evolution with detailed feedback and chemical enrichment. For the first time on galaxy scales, we simultaneously model multi-channel stellar feedback in detail, 
%mm3 accounting for stellar winds, 
     using individual star particles to model
core collapse and Type Ia SNe, ionizing radiation followed through radiative transfer, photoelectric heating, Lyman-Werner radiation
%mm3 using individual star particles
    and pollution from AGB and massive stellar winds. This
%mm3
    treatment of feedback, 
coupled with the detailed chemistry and heating/cooling physics followed with \textsc{Grackle}, allows us to capture realistic galaxy evolution in detail. In this work, we apply these methods to simulate the evolution of an isolated, low-mass, dwarf galaxy modeled after the $z=0$ properties of the Leo P ultrafaint dwarf galaxy. We present an overview of the properties of this simulation 
in this work.
% AE: mention companion paper here if publish at same time

For our simulated dwarf galaxy, we find:
\begin{enumerate}
\item Multi-channel feedback is effective in this galaxy in regulating star formation to occur at a rate consistent with the Kennicutt-Schmidt relationship and the extended Schmidt law in observed galaxies. See Figures~\ref{fig:sfr_mass_evolution} and \ref{fig:KS}.

@@

\item 
%mm3 A thin, yet resolved, gas scale height ($\sim 50$~pc)
   The gas scale height is thin ($\sim 50$~pc), yet resolved, 
beyond our stellar radius, with larger scale heights ($\sim 100$--200~pc) driven by feedback interior to the stellar radius. This is comparable to the resolution limit of the diffuse \ion{H}{1} in observed, gaseous ultrafaint dwarfs. (See Figure~\ref{fig:scale_height}).

\item The ISRF of our galaxy varies strongly in time, by orders of magnitude, though
%mm3 it is relatively uniform in space at any given time. [this is contradicted by Figure \ref{fig:ISRF_2D}]
    it has a relatively uniform radial profile averaged over tens of megayears.
It is unclear how important these fluctuations are as a source of feedback, or if the affect can be approximated with a time-averaged radial profile. (See Figures \ref{fig:ISRF} and \ref{fig:ISRF_2D}.)

\item 
%mm3 (\textbf{Possibly remove depending on follow up}) Formation of a significant amount of H$_2$ 
     A significant amount of H$_2$ forms
through gas phase reactions with H$^-$. This dense gas is more resistive to feedback driven outflows than the less dense gas in the galaxy. This leads to high molecular gas fractions, with an H$_2$ mass comparable to that of \ion{H}{1} by the end of the simulation. This result requires further investigation to determine if this behavior is universally true of ultrafaint dwarf galaxies. See Figure~\ref{fig:ISM_evolution}.
